### 2016-10-04 Projects/DelCo_2012/Report/DelCo_2012.tex
Added majority of Nelson's edits to CAF report

\vspace*{2\baselineskip}
August 2016}

\vspace*{2\baselineskip}
{\em $^+$\small Currently with JENSEN HUGHES}
\\
{\em $^*$\small Currently with UL FSRI}
}

@@

\large
\flushright{\bf National Institute of Standards and Technology Technical Note 1927 \\
Natl.~Inst.~Stand.~Technol.~Tech.~Note~1927, \pageref{LastPage} pages (August 2016) \\
CODEN: NTNOEF }

@@

\begin{enumerate}
\item  Perform a literature review of available published information on the use of CAF for structural fire fighting~\cite{Mitchell:1}
\item  Conduct a workshop that brings interested and experienced parties together to discuss CAF experience and to provide guidance on the proposed experimental research plan~\cite{Grant:2011} 
\item  Conduct and document well-instrumented large-scale enclosure fire tests using both CAF and plain water to evaluate the effectiveness of CAF for interior structural fire fighting
\item  Conduct and document identified fire ground related experiments to evaluate differences between CAF and plain water hose lines related to issues such as effective hose stream throw and distribution, nozzle reaction force, the force to kink the hose, potential for slipping on a variety of surfaces, and flow separation~\cite{Carracino:2013,Dicus:2013,LaPolla:2012}. 
\end{enumerate}

@@

Based on the results of the NIST 1998 study, the fire fighting agents, commonly referred to as Class A foams, were superior to water in reducing contact angle; cooling and penetrating wood-based fuels; mass retention on the wood fuels; and ignition inhibition of the wood fuels.  When applied in the form of foam, the water agent solutions provide enhanced radiation exposure protection to a structure relative to plain water. The foams enhanced the water by keeping the water in place on vertical surfaces such as walls and the reduced inter-facial surface tension allowed the water to penetrate absorbent fuels. These enhancements also make the agents valuable tools for mop-up. The other experiments conducted during this study---specific heat, drop size, wood crib suppression, smoke generation, and Class D fire suppression---yielded similar results for both Class A foams and plain water.

@@

Montgomery County (MD) Fire and Rescue Service (MCFRS) is one of the larger fire departments in Maryland and was a committed user of CAF with 35 CAFS front line engines. MCFRS had experienced two incidents where firefighters received burns in residential structure fires while CAF was being used to fight the fire during an interior attack. Concerns were raised within the department regarding the ability of CAF to cool fire gases and absorb heat as well as plain water. As a result, on March 4, 2010, the Fire Chief of MCFRS issued a General Order prohibiting the use of CAF during offensive interior attacks. In this same order, the Chief directed the Division Chief of Operations to continue to evaluate the use of CAF during structure fires and to report back his findings and recommendations. That order was the catalyst for this study.

@@

All of the experiments were conducted at full scale in compartments and structures that were representative of room sizes in a residential occupancy. Technical panel members whose departments routinely use CAF for structural fire fighting were invited and present at many of the fire suppression experiments. The experiments were planned during the workshop.  Three types of experiments examining spray density measures, gas layer cooling, and fire suppression were conducted in the fall of 2012. The data from these experiments were presented to the technical panel members. The fire suppression experiments were conducted with a fuel package that included upholstered furniture. After a review of the data, a decision was made to conduct an additional series of suppression experiments with a fuel package that was composed mainly of wood-based materials. The thinking was that the synthetic materials in the upholstered furniture burned only on the surface and were easily extinguished. wood-based fuels would char and present a deep seated fire challenge to the extinguishing agents.  After the wood-based fuel package experiments were completed in 2013, it was decided in discussions with members of the technical panel that other experimental conditions needed to be examined: room size, additional ventilation, and a manually operated handline. The issue of a larger fire compartment had also been raised during the workshop discussion, but the desire for replicate experiments and financial limitations removed the large floor plan experiments from the original plan. To accomplish the last series of experiments, additional funding was obtained from NIST, and the experiments were conducted in 2015. The fuel load contained both wood-based wall and ceiling finishes, pallets, carpeting, and upholstered furniture.

@@

\section{Description of Experiments}
\label{sec:desc_experiments}
The experimental setup for this test series is described by the three types of experiments conducted: spray density measures, gas layer cooling, and fire suppression.  Baseline operating conditions were agreed upon at the CAFS workshop for these experiments (Table~\ref{tab:op_condition})~\cite{Grant:2011}. The nozzles chosen for these experiments were Task Force Tips (TFT) Metro 1, fixed gallonage nozzles with a flow rate of 120~gpm. These are automatic nozzles which have a 7/8~inch slug-tip and can be varied from a straight stream to a wide fog stream for application. The setup also included 100~ft of 1-3/4~inch hose. Nozzles were mounted to a ``blitz-fire'' monitor device and placed inside the structures. The nozzles were controlled and operated via a gate valve from the exterior. The exterior control had a gated wye to allow the flow of CAF through a hose directed away from the test structure. When the flow of CAF was needed for the experiment, the gated wye would be switched to allow the CAF to flow into the test structure. This was done to minimize the build-up of ``dry'' foam or a slug upon the initial flow of the nozzle into the test structure.

\begin{table}[!ht]
\centering
\captionof{table}{Operating conditions for experiments}\label{tab:op_condition}
\begin{tabular}{ll}
\toprule[1.5pt]
Equipment    &   Operating Condition \\
\midrule
Nozzle       & Metro 1, fixed gallonage fog, straight, or 7/8~inch solid stream \\
Hoseline     & 100~ft of 1-3/4~inch hoseline \\
Foam         & 0.3\% foam concentration \\
Pressure     & Water and Air pressure of 100 - 120~psi \\
\bottomrule[1.25pt]
\end{tabular}\par
\end{table}

@@

The interior walls of the burn room were lined with 13~mm (0.5~in) thick cement board. The ceiling of the burn room was the exposed ``floor assembly". The walls of the hallway and entry foyer were composed of 16~mm (0.625~in) Type X gypsum board. The ceiling of the hallway and entry foyer was composed of two layers of 13~mm (0.5~in) thick cement board.

@@

The outer walls of the bottom floor of the two-story structure were identical to that of the single-story structures described above but with an open floor plan. The interior dimensions were 6.1~m (20~ft) wide, 11~m (36~ft) long and 2.4~m (8~ft) high. Figure~\ref{fig:dimensioned_first_2story} shows the dimensions of the reconfigured first floor. A stairwell was built to connect the floors. Stairs started 1.6~m (5.25~ft) off the south wall with a width of 1.2~m (4~ft) off the east wall. The second story walls were wood framed with 51~mm (2~in) by 102~mm (4~in) studs. The studs were set to 0.40~m (16~in) centers. The interior  walls were protected by 16~mm (0.63~in) fire rated gypsum board, 16~mm (0.63~in) durarock board, and a second layer of 16~mm (0.63~in) fire rated gypsum board. The exterior walls were protected with 11~mm (0.31~in) oriented strand board and 8~mm (0.31~in) fiber cement lap siding. Figure~\ref{fig:dimensioned_second_2story} shows the dimensions of the second story of the modified west structure.

@@

The structures were instrumented for temperature, gas velocity, and heat flux measurements. Gas temperatures in the burn rooms were measured with bare-bead, Chromel-Alumel (type K) thermocouples. Additional single, sheathed thermocouples were installed in conjunction with the bi-directional probes for gas velocity measurements. The single thermocouples were bare-bead, Chromel-Alumel (type K) thermocouples with a 1.0~mm (0.04~in) nominal diameter. The thermocouple was protected with an 3.2~mm (0.125~in) diameter inconel sheath. The sheathing protects the thermocouple from water but slows the response time because of the additional thermal mass. Schmidt-Boelter gauges were used to measure both total heat flux and radiant heat flux (radiometer). A radiometer is a total heat flux gauge with a zirconium plate to prevent contributions from convective heat transfer. A legend, which clarifies the instrumentation schematics discussed in the following sections, is included in Figure~\ref{fig:Instrumentation_Legend}.

@@

Table~\ref{tab:Test_Results_2} provides the impact of suppression for the water, class A foam, and CAF tests conducted in the two-story structure. The measurement location was on the floor above the fire, at the doorway to the stairwell which connects the floors. The thermocouples at this measurement location were isolated from any direct contact from the suppression stream. The table provides the change in temperature 0.3~m (1~ft) below the ceiling and 0.3~m (1~ft) above the floor as well as the change in heat flux 1~m (3~ft) above the floor on the 2nd floor level in front of the stairway door. Figure~\ref{fig:tcs_hf_second_2story} shows the position of the thermocouple array (TC A10) as well as the three pairs of heat flux gauges. Table~\ref{tab:Test_Results_2} shows the heat flux at position 1, that faces in the vertical direction (toward the ceiling). Note that these thermocouples are inconel shielded thermocouples and have slower response times (smaller $\Delta$T's over similar time intervals) compared to the bare-bead thermocouples used in the single-story tests. The values reported in the table are immediately before and after the application period to assess the application agent, stream, and location on interior conditions.

@@

As mentioned in the results section, the conditions at TC Array 3 represented the ``best'' gas cooling results because due to the array's position, it was impacted the least by agents hitting and cooling the thermocouples themselves. Under the given conditions of a residential scale room and the amount of water introduced into the hot gas layer, it should be no surprise that the water is doing the work. Air filled bubbles in the CAF stream had no discernible impact, either positive or negative, in the gas cooling experiments. At the same time, there was no measurable negative effect of the air bubbles with regard to gas cooling.

@@

The data and results from this test series do not support the premise of firefighters being injured during an interior fire fighting operation with CAF because CAF provided cooling similar to that of the equivalent flow rate of plain water.

@@

The fire rooms in these experiments ranged in floor area from approximately 27~m$^2$ (291~ft$^2$) to 51.4~m$^2$ (553~ft$^2$). Applying the NFA formula based on floor area yields a required flow rate for the single-story fire room of 369~l/min (97~gpm). The required flow rate for the fire room in the lower level of the two-story would be approximately 700~l/min (184~gpm). Comparing the hose stream flow rates to the NFA Formula requirements indicates that 454~l/min (120~gpm) was sufficient for the smaller room fire in the single-story but was below the required flow rate of approximately 700~l/min (184~gpm) for the large fire compartment in the two-story structure. This is not consistent with the observations of a rapid fire knockdown with both water and CAF hose streams flowing 454~l/min (120~gpm).

@@

During the workshop conducted for this project~\cite{Grant:2011}, there were discussions that CAF could suppress, or knockdown, a fire three to five times faster than water. During the course of this study, there were no significant differences demonstrated between CAF and water for use in interior structural fire fighting on a residential scale. Going back to the premise of the study, preventing firefighters from being injured during interior fire fighting operations with CAF, there is nothing in these fire suppression results to suggest that CAF would present a thermal hazard, provided that an adequate amount of water is being flowed for the given fire condition.

@@

One of the challenges with suppression research or determining the required flow rate of water needed on a building fire is the lack of data on realistic baseline requirements of the total amount of water needed for gas cooling and suppression. In this study, it seems clear that there was a sufficient amount of water for gas cooling and suppression. One question that arises is, how much could the flow rate of water have been reduced before a significant difference in performance was observed?

Information from actual incidents in the field shaped this study. From the gas cooling and suppression perspective, the results do not indicate that the use of CAF would result in a higher thermal hazard. However, there are other issues like nozzle reaction force and flow separation which have been covered in the Cal Poly fire ground studies that can be addressed in training~\cite{Carracino:2013,Dicus:2013,LaPolla:2012}.  It is important to provide adequate training when adopting a new method or new technology. Perhaps these types of issues should be examined further to better understand how suppression activities take place in the field.

Additional research is needed to examine the impact of nozzle type and nozzle movement on the fire environment. It is possible that some of the discomfort or injuries that firefighters experience during interior fire fighting could be the result of a poor choice of nozzle or nozzle movement that results in displacing hotter gases into the area where the firefighter is located. Currently the fire service relies on experience to develop a feel for the correct combination of nozzle type and movement. A study may be able to demonstrate a clear cause and effect relationship between the amount of water flowing, the type of nozzle, and the movement of the nozzle and how such choices can impact the flow of gases within the structure. Based on the results of this study, it appears that training and understanding the impact of nozzle movement are ways to look for a more effective use of water.

@@

Full-scale fire experiments designed to address those questions were conducted over a period of years from 2012 through 2015. The experimental configurations were designed to be representative of residential fire scenarios. The water and CAF flows were based on those used by a fire department with 35 CAFS front line engines. This same department also experienced injuries to firefighters while using CAF during interior fire fighting operations in residential structures. This study was composed of three types of experiments: spray density, gas cooling, and fire suppression. In each type of experiment, a comparison was conducted between plain water and CAF being discharged from nozzles of similar types with similar mass flow rates of water. The flow rate for each water experiment was 120~gpm. For the CAF experiments, 60~cfm of air and a 0.3~\% flow of Class A foam concentrate were added to the water flow.

>>>>

\vspace*{2\baselineskip}
October 2016}

\vspace*{2\baselineskip}
{\em $^+$\small Currently with JENSEN HUGHES, Baltimore MD}
\\
{\em $^*$\small Currently with Underwriters Laboratories Firefighter \\ 
Safety Research Institute, Columbia MD}
}

@@

\large
\flushright{\bf National Institute of Standards and Technology Technical Note 1927 \\
Natl.~Inst.~Stand.~Technol.~Tech.~Note~1927, \pageref{LastPage} pages (October 2016) \\
CODEN: NTNOEF }

@@

\begin{enumerate}
\item  Perform a literature review of available published information on the use of CAF for structural fire fighting~\cite{Mitchell:1}
\item  Conduct a workshop that brings interested and experienced parties together to discuss CAF experience and to provide input on the proposed experimental research plan~\cite{Grant:2011} 
\item  Conduct and document well-instrumented large-scale enclosure fire tests using both CAF and plain water to evaluate the effectiveness of CAF for interior structural fire fighting
\item  Conduct and document identified fire ground related experiments to evaluate differences between CAF and plain water hose lines related to issues such as effective hose stream throw and distribution, nozzle reaction force, the force to kink the hose, potential for slipping on a variety of surfaces, and flow separation~\cite{Carracino:2013,Dicus:2013,LaPolla:2012}. 
\end{enumerate}

@@

Based on the results of the NIST 1998 study, the fire fighting agents, commonly referred to as Class A foams, were superior to water in reducing contact angle; cooling and penetrating wood-based fuels; mass retention on the wood fuels; and ignition inhibition of the wood fuels.  When applied in the form of foam, the water agent solutions provide enhanced radiation exposure protection to a structure relative to plain water. The foams enhanced the water by keeping the water in place on vertical surfaces such as walls and the reduced inter-facial surface tension allowed the water to penetrate adsorbent fuels. These enhancements also make the agents valuable tools for mop-up or overhaul. The other experiments conducted during this study---specific heat, drop size, wood crib suppression, smoke generation, and Class D fire suppression---yielded similar results for both Class A foams and plain water.

@@

Montgomery County (MD) Fire and Rescue Service (MCFRS) is one of the larger fire departments in Maryland and was a committed user of CAF with 35 CAFS front line engines. MCFRS had experienced two incidents where firefighters received burns in residential structure fires while CAF was being used to fight the fire during an interior attack. Concerns were raised within the department regarding the ability of CAF to cool fire gases and absorb heat as well as plain water. As a result, on March 4, 2010, the Fire Chief of MCFRS issued a General Order prohibiting the use of CAF during offensive interior attacks. In this same order, the Chief directed the Division Chief of Operations to continue to evaluate the use of CAF during structure fires and to report back his findings and recommendations. That order was a catalyst for this study.

@@

All of the experiments were conducted at full scale in compartments and structures that were representative of room sizes in a residential occupancy. Technical panel members whose departments routinely use CAF for structural fire fighting were invited and present at many of the fire suppression experiments. The experiments were planned during the workshop.  Three types of experiments examining spray density measures, gas layer cooling, and fire suppression were conducted in the fall of 2012. The data from these experiments were presented to the technical panel members. The fire suppression experiments were conducted with a fuel package that included upholstered furniture. After a review of the data, a decision was made to conduct an additional series of suppression experiments with a fuel package that was composed mainly of wood-based materials. The logic was that the synthetic materials in the upholstered furniture burned only on the surface and were easily extinguished. Wood-based fuels would char and present a deep seated fire challenge to the extinguishing agents.  After the wood-based fuel package experiments were completed in 2013, it was decided in discussions with members of the technical panel that other experimental conditions needed to be examined: room size, additional ventilation, and a manually operated handline. The issue of a larger fire compartment had also been raised during the workshop discussion, but the desire for replicate experiments and financial limitations removed the large floor plan experiments from the original plan. To accomplish the last series of experiments, additional funding was obtained from NIST, and the experiments were conducted in 2015. The fuel load contained both wood-based wall and ceiling finishes, pallets, carpeting, and upholstered furniture.

@@

\section{Description of Experiments}
\label{sec:desc_experiments}
The experimental setup for this test series is described by the three types of experiments conducted: spray density measures, gas layer cooling, and fire suppression.  Baseline operating conditions were agreed upon at the CAFS workshop for these experiments (Table~\ref{tab:op_condition})~\cite{Grant:2011}. The nozzles chosen for these experiments were Task Force Tips (TFT) Metro 1, fixed gallonage nozzles with a flow rate of 120~gpm. These are automatic nozzles which have a 7/8~inch (22~mm) slug-tip and can be varied from a straight stream to a wide fog stream for application. The setup also included 100~ft (30.5~m) of 1-3/4~inch (44~mm) hose. Nozzles were mounted to a ``blitz-fire'' monitor device and placed inside the structures. The nozzles were controlled and operated via a gate valve from the exterior. The exterior control had a gated wye to allow the flow of CAF through a hose directed away from the test structure. When the flow of CAF was needed for the experiment, the gated wye would be switched to allow the CAF to flow into the test structure. This was done to minimize the build-up of ``dry'' foam or a slug upon the initial flow of the nozzle into the test structure.

\begin{table}[!ht]
\centering
\captionof{table}{Operating conditions for experiments}\label{tab:op_condition}
\begin{tabular}{ll}
\toprule[1.5pt]
Equipment    &   Operating Condition \\
\midrule
Nozzle       & Metro 1, fixed gallonage fog, straight, or 7/8~inch solid stream \\
Hoseline     & 100~ft of 1-3/4~inch hoseline \\
Foam         & 0.3\% foam concentration \\
Pressure     & Water and Air pressure of 100--120~psi (689--827~kPa)\\
\bottomrule[1.25pt]
\end{tabular}\par
\end{table}

@@

The interior walls of the burn room were lined with 13~mm (0.5~in) thick cement board. The ceiling of the burn room was the exposed ``floor assembly''. The walls of the hallway and entry foyer were composed of 16~mm (0.625~in) Type X gypsum board. The ceiling of the hallway and entry foyer was composed of two layers of 13~mm (0.5~in) thick cement board.

@@

The outer walls of the bottom floor of the two-story structure were identical to that of the single-story structures described above but with an open floor plan. The interior dimensions were 6.1~m (20~ft) wide, 11~m (36~ft) long and 2.4~m (8~ft) high. Figure~\ref{fig:dimensioned_first_2story} shows the dimensions of the reconfigured first floor. A stairwell was built to connect the floors. Stairs started 1.6~m (5.25~ft) off the south wall with a width of 1.2~m (4~ft) off the east wall. The second story walls were wood framed with 51~mm (2~in) by 102~mm (4~in) studs. The studs were set to 0.40~m (16~in) centers. The interior  walls were protected by 16~mm (0.63~in) fire rated gypsum board, 16~mm (0.63~in) Durock board, and a second layer of 16~mm (0.63~in) fire rated gypsum board. The exterior walls were protected with 11~mm (0.31~in) oriented strand board and 8~mm (0.31~in) fiber cement lap siding. Figure~\ref{fig:dimensioned_second_2story} shows the dimensions of the second story of the modified west structure.

@@

The structures were instrumented for gas temperature, gas velocity, and heat flux measurements. Gas temperatures in the burn rooms were measured with bare-bead, Chromel-Alumel (type K) thermocouples. Additional single, sheathed thermocouples were installed in conjunction with the bi-directional probes for gas velocity measurements. The single thermocouples were bare-bead, Chromel-Alumel (type K) thermocouples with a 1.0~mm (0.04~in) nominal diameter. The thermocouple was protected with an 3.2~mm (0.125~in) diameter inconel sheath. The sheathing protects the thermocouple from water but slows the response time because of the additional thermal mass. Schmidt-Boelter gauges were used to measure both total heat flux and radiant heat flux (radiometer). A radiometer is a total heat flux gauge with a zirconium plate to prevent contributions from convective heat transfer. A legend, which clarifies the instrumentation schematics discussed in the following sections, is included in Figure~\ref{fig:Instrumentation_Legend}.

@@

Table~\ref{tab:Test_Results_2} provides the impact of suppression for the water, class A foam, and CAF tests conducted in the two-story structure. The measurement location was on the floor above the fire, at the doorway to the stairwell which connects the floors. The thermocouples at this measurement location were isolated from any direct contact from the suppression stream. The table provides the change in temperature 0.3~m (1~ft) below the ceiling and 0.3~m (1~ft) above the floor as well as the change in heat flux 1~m (3~ft) above the floor on the second floor level in front of the stairway door. Figure~\ref{fig:tcs_hf_second_2story} shows the position of the thermocouple array (TC A10) as well as the three pairs of heat flux gauges. Table~\ref{tab:Test_Results_2} shows the heat flux at position 1, that faces in the vertical direction (toward the ceiling). Note that these thermocouples are inconel shielded thermocouples and have slower response times (smaller $\Delta$T's over similar time intervals) compared to the bare-bead thermocouples used in the single-story tests. The values reported in the table are immediately before and after the application period to assess the application agent, stream, and location on interior conditions.

@@

As mentioned in the results section, the conditions at TC Array 3 represented the ``best'' gas cooling results because due to the array's position, it was impacted the least by agents hitting and cooling the thermocouples themselves. Under the given conditions of a residential scale room and the amount of water introduced into the hot gas layer, it should be no surprise that the water is absorbing the energy from the upper layer. Air filled bubbles in the CAF stream had no discernible impact, either positive or negative, in the gas cooling experiments. At the same time, there was no measurable negative effect of the air bubbles with regard to gas cooling.

@@

The data and results from this test series are not consistent with a hypothesis that firefighters were being injured during an interior fire fighting operation with CAF because CAF provided cooling similar to that of the equivalent flow rate of plain water.

@@

The fire rooms in these experiments ranged in floor area from approximately 27~m$^2$ (291~ft$^2$) to 51.4~m$^2$ (553~ft$^2$). Applying the NFA formula based on floor area yields a required flow rate for the single-story fire room of 369~l/min (97~gpm). The required flow rate for the fire room in the lower level of the two-story would be approximately 700~l/min (184~gpm). Comparing the hose stream flow rates to the NFA formula requirements indicates that 454~l/min (120~gpm) was sufficient for the smaller room fire in the single-story but was below the required flow rate of approximately 700~l/min (184~gpm) for the large fire compartment in the two-story structure. This is not consistent with the observations of a rapid fire knockdown with both water and CAF hose streams flowing 454~l/min (120~gpm).

@@

During the workshop conducted for this project~\cite{Grant:2011}, it was suggested that CAF could suppress, or knockdown, a fire three to five times faster than water. The data collected during this study does not demonstrate significant differences between CAF and water for use in interior structural fire fighting on a residential scale. Going back to the premise of the study, preventing firefighters from being injured during interior fire fighting operations with CAF, there is nothing in these fire suppression results to suggest that CAF would present a thermal hazard, provided that an adequate amount of water is being flowed for the given fire condition.

@@

One of the challenges with suppression research or determining the required flow rate of water needed on a building fire is the lack of data on realistic baseline requirements of the total amount of water needed for gas cooling and suppression. In this study, the data demonstrates that there was a sufficient amount of water for gas cooling and suppression. One question that arises is, how much could the flow rate of water have been reduced before a significant difference in performance was observed?

Information from actual incidents in the field shaped this study. From the gas cooling and suppression perspective, the results did not demonstrate significant differences between plain water and CAF. However, there are other issues like nozzle reaction force and flow separation which have been covered in the Cal Poly fire ground studies that can be addressed in training~\cite{Carracino:2013,Dicus:2013,LaPolla:2012}.  It is important to provide adequate training when adopting a new method or new technology. These types of issues should be examined further to better understand how suppression activities take place in the field.

Additional research is needed to examine the impact of nozzle type and nozzle movement on the fire environment. It is possible that some of the discomfort or injuries that firefighters experience during interior fire fighting could be the result of a poor choice of nozzle or nozzle movement that results in displacing hotter gases into the area where the firefighter is located. Currently, the fire service relies on experience to develop a feel for the correct combination of nozzle type and movement. A study may be able to demonstrate a clear cause and effect relationship between the amount of water flowing, the type of nozzle, and the movement of the nozzle and how such choices can impact the flow of gases within the structure. Based on the results of this study, it appears that training and understanding the impact of nozzle movement may be potential paths to more effective use of water.

@@

Full-scale fire experiments designed to address those questions were conducted over a period of years from 2012 through 2015. The experimental configurations were designed to be representative of residential fire scenarios. %The water and CAF flows were based on those used by a fire department with 35 CAFS front line engines. This same department also experienced injuries to firefighters while using CAF during interior fire fighting operations in residential structures. 
This study was composed of three types of experiments: spray density, gas cooling, and fire suppression. In each type of experiment, a comparison was conducted between plain water and CAF being discharged from nozzles of similar types with similar mass flow rates of water. The flow rate for each water experiment was 120~gpm. For the CAF experiments, 60~cfm of air and a 0.3~\% flow of Class A foam concentrate were added to the water flow.
